They're Clutching up! Team Momentum in Round-Based Esports

### 2022-08-19 paper/paper.Rmd
ALL THE SYMBOL CHANGES

There have been a handful of studies of the distribution of games played in a series of a professional sport. Most assume a constant probability $p$ of a given team winning a game in the series, regardless of the series state. Mosteller [-@mosteller1952] observed that the American League had dominated the National League in Major League Baseball's (MLB) World Series matchups, implying that matchups should not modeled with $p = 0.5$. Mosteller proposed three approaches for identifying the optimal constant probability value of the stronger team in the World Series, finding $p \approx 0.65$. in each case: (1) solving for $p$ from the empirical average number of games won by the loser of the series, which he called the "method of moments" approach; (2) maximizing the likelihood that the sample would have been drawn from a population in which the probability of a team winning a game is constant across the series (i.e. maximum likelihood), and (3) minimizing the chi-square goodness of fit statistic for $p$.

Chance [-@chance2020] re-examines the constant probability notion in Major League Baseball's World Series (1923--2018), the National Basketball Association's Finals (1951--2018), and the National Hockey League's Stanley Cup (1939--2018). Chance finds strong evidence against the null hypothesis of $p = 0.5$ in the MLB and NHL championship series when applying Mosteller's first and second methods.[\^8] [\^8]: Chance goes on to outline a conditional probability framework (likelihood of winning a game given the series state) which can exactly explain the distribution of the number of games played.

@@

Studying both player streaks and team streaks in basketball, in both observational and controlled settings. Gilovich et al. [-@gilovich1985] do not find evidence for the hot hand phenomenon. However, Miller and Sanjurjo [-@miller2018] refuted the conclusions of Gilovich et al., proving a framework for quantifying streak selection bias, which effectively works in the manner posited by the gambler's fallacy[^9]. Specifically, Miller and Sanjurjo say that a "bias exists in a common measure of the conditional dependence of present outcomes on streaks of past outcomes in sequential data" implying that, under i.i.d. conditions, "the proportion of successes among the outcomes that immediately follow a streak of consecutive successes is expected to be strictly less than the underlying (conditional) probability of success". When applying streak selection bias to Giloveih et al.'s data, Miller and Sanjurjo come to the opposite conclusions as Giloveih et al.

[^9]: Streak selection bias, "in conjunction with a quasi-Bayesian model of decision making under sample size neglect... provides a novel structural candidate explanationfor the persistence of gambler's fallacy beliefs", per Miller and Sanjurjo [-@miller2018].

Despite the plethora of existing research on games played in a series and momentum in sports, these topics have yet to be investigated heavily in esports. Work has been done to examine in-round win probability in other FPS titles such as Counter-Strike [@xenopoulos2022] and Valorant [@derover2021], both of which are round-based like CoD SnD. However, research on round-level trends is sparse, perhaps because games like Counter-Strike and Valorant both have economic aspects that can create clear advantages on side in a given round, given how prior rounds played out.[^10]

[^10]: Additionally, both Counter-Strike and Valorant have overtime rules and blocked offensive/defensive roles (i.e. playing either offense or defense for many consecutive rounds).

@@

The general formula for the probability $P_E(r)$ that a best-of-$s$ series ($s = 11$ for CoD SnD) lasts $r$ rounds given constant probability $p$ of one team[^11] winning each round is

[^11]: If $p > 0.5$, then one might say that this team is the better team (known in hindsight).

@@

```{=tex}
\begin{equation}\protect\hypertarget{eq:series-length}{}{
P_E(r) = \frac{(r - 1)!}{(m - 1)!(r - s)!}(p^{m}(1 - p)^{r - m} + p^{r - m}(1 - p)^m).
}\label{eq:series-length}\end{equation}
```
The chi-square goodness of fit statistic can be used to test the constant found probability hypothesis for $0 \leq p \leq 1$.

```{=tex}
\begin{equation}\protect\hypertarget{eq:chi-squ}{}{
\chi^2 = \sum^R \frac{(P_O(r) - P_E(r))^2}{P_E(r)}
}\label{eq:chi-squ}\end{equation}
```
For CoD SnD, $r \in R [6, 7, 8, 9, 10, 11]$.

@@

Without accounting for the streak selection bias described by Miller and Sanjurjo [-@miller2018], the round win probability after a streak of $k$ round wins for a best-of-$s$ series ($s = 11$ for CoD SnD), accounting for the series lasting $r$ rounds known in hindsight, is

```{=tex}
\begin{equation}\protect\hypertarget{eq:pwkr}{}{
  p^{w|kr} = p(\text{win | k, r})=\{
    \begin{array}{cl}
    \frac{m - k}{i - k}, & \text{team wins series} \\
    \frac{s - m - k}{i - k}, & \text{team loses series}
    \end{array}.
  }
\label{eq:pwkr}
\end{equation}
```
The "hindsight" round loss probability after a streak of $k$ round losses for a series lasting $r$ rounds, $p^{l|kr}$, can be formulated symmetrically.[^12]

[^12]: Note that these probabilities are separate from the constant round probabilities in Sections \ref{sec:method-rounds-played} and \ref{sec:results-rounds-played}.

However, as shown by Miller and Sanjurjo, we should expect the observed proportion of rounds wins after a streak $k$ rounds wins for a series lasting $r$ rounds, $\hat{P}^{w|kr}$, to be strictly less than $p^{w|kr}$.[^13]

[^13]: See Appendix A of Miller and Sanjurjo [-@miller2018] for the proof.

@@

Miller and Sanjurjo [-@miller2018] note that there does not exists a closed form representation for $E[\hat{P}_k(\mathbf{X}) | I_k(\mathbf{X}) \neq \emptyset]$ ($\hat{P}^{w|kr}$ for CoD SnD) for $k > 1$. Figure \ref{fig:npk} shows this.[^14]

[^14]: I've adapted the code from Vafa [-@vafa2017].

![The expected value](images/npk_replicated.png)

The expected value of the proportion of successes on trials that immediately follow k consecutive successes, Pˆk(X), as a function of the total number of trials n,for different values of k and probabilities of success p,

@@

First, I investigate the constant probability assumption and the distribution of rounds played in a series. Chance's [-@chance2020] work is closely related to mine, and, in fact, provides a guide for this investigation.

@@

Using Equation \ref{eq:chi-squ}, $\chi^2 = 16.0$ (p-value of 0.0068) for a naive constant round win probability $p_0 = 0.5$. Thus, I can comfortably reject the constant probability null hypothesis for $p_0 = 0.5$, even at a confidence level of $\alpha = 0.01$. Afterwards, I investigate post-streak win rates, building on my learnings from the constant probability assumption analysis. I agree with Miller and Sanjurjo's [-@miller2018] framework for streak selection bias, accounting for it in when analyzing momentum.

```{r}
#| label: cod_series_outcome_prop
.max_round <- 11
.cutoff <- 6
prob_of_series_lasting_r_rounds <- function(r, p = 0.5, s = .max_round) {
  m <- as.integer((s + 1) / 2)
  (factorial(r - 1) / (factorial(m - 1) * factorial(r - s))) * (p^m * (1 - p)^(r - m) + p^(r - m) * (1 - p)^m)
}

@@

  expected_round_streaks <- tibble(
    n_rounds = .cutoff:.max_round
  ) |> 
    mutate(
      series_prop = map_dbl(n_rounds, ~prob_of_series_lasting_r_rounds(.x, max_round = .max_round, p = !!p)),
      n_expected_series_streaks = map_dbl(n_rounds, ~expected_series_streaks_of_outcomes(.cutoff, .x - .cutoff))
    ) |> 
    transmute(
      n_rounds,
      series_prop,
      prop = series_prop / n_expected_series_streaks
    )

@@

```{r}
#| label: cod_series_outcome_prop-table
cod_streaks_naive_res$series |>
  select(
    n_rounds,
    prop_expected,
    prop_actual,
    n_actual
  ) |>
  arrange(n_rounds) |>
  gt::gt() |>
  gt::cols_label(
    n_rounds = 'r',
    prop_expected = 'P_{E}(r)',
    prop_actual = 'P_{O}(r)',
    n_actual = 'N_O'
  ) |>
  gt::fmt_percent(
    decimals = 1,
    columns = c('prop_expected', 'prop_actual')
  ) |>
  gt::tab_spanner(
    label = 'Observed',
    columns = c('prop_actual', 'n_actual')
  ) |>
  gt::as_latex()
```

Table \ref{tbl:cod-prob-series-lasting-r-rounds} shows the expected probabilities for $s = 11$ under the naive assumption that $p_0 = 0.5$ for a given round, along with the observed proportions $P_O(r)$ in CoD SnD.

```{=tex}
\begin{longtable}{rrrr}
\caption{The probabilities that a best-of-11 series lasts $r$ rounds ($P_E(r)$, where $r \in [6, 7, 8, 9, 10, 11]$) under the assumption that each team has a 50\% probability ($p_0 = 0.5$) of winning each game. The observed frequencies observed are shown as a count $N_O(r)$ and as a proportion $P_O(r)$ of all series.}\label{tbl:cod-prob-series-lasting-r-rounds} \\
\toprule
&  & \multicolumn{2}{c}{Observed} \\ 
\cmidrule(lr){3-4}
$r$ & $P_E(r)$ & $P_O(r)$ & $N_O(r)$ \\ 
\midrule
6 & $3.1\%$ & $4.7\%$ & 40 \\ 
7 & $9.4\%$ & $11.9\%$ & 101 \\ 
8 & $16.4\%$ & $16.5\%$ & 141 \\ 
9 & $21.9\%$ & $21.7\%$ & 185 \\ 
10 & $24.6\%$ & $21.5\%$ & 183 \\ 
11 & $24.6\%$ & $23.7\%$ & 202 \\ 
\bottomrule
\end{longtable}
```
```{r}
#| label: cod_series_outcomes_naive_chi
generate_chi_label <- function(statistic, p.value) {
  sprintf('%.1f (%s)', statistic, ifelse(p.value <= 0.01, '<=0.01', as.character(round(p.value, 2))))
}

@@

```{r}
#| label: cod-alternative-to-p=0.5-method-1
## p. 365 on https://math.mit.edu/classes/18.095/2016IAP/lec9/Sports_Mosteller1952_WorldSeries.pdf
theoretical_cod_series_length <- function(p) {
  6 * prob_of_series_lasting_r_rounds(n = 6, p = p, max_round = 11) +
    7 * prob_of_series_lasting_r_rounds(n = 7, p = p, max_round = 11) +
    8 * prob_of_series_lasting_r_rounds(n = 8, p = p, max_round = 11) +
    9 * prob_of_series_lasting_r_rounds(n = 9, p = p, max_round = 11) +
    10 * prob_of_series_lasting_r_rounds(n = 10, p = p, max_round = 11) +
    11 * prob_of_series_lasting_r_rounds(n = 11, p = p, max_round = 11)
}
cod_rounds_per_series <- cod_streaks_naive_res$series |>
  summarize(
    actual = sum(n_rounds * prop_actual),
    expected = sum(n_rounds * prop_expected)
  )

@@

maximize_cod_series_p <- function(p) {
  prob_of_series_lasting_r_rounds(n = 6, p = p, max_round = 11)^cod_n6adj *
    prob_of_series_lasting_r_rounds(n = 7, p = p, max_round = 11)^cod_n7adj *
    prob_of_series_lasting_r_rounds(n = 8, p = p, max_round = 11)^cod_n8adj *
    prob_of_series_lasting_r_rounds(n = 9, p = p, max_round = 11)^cod_n9adj *
    prob_of_series_lasting_r_rounds(n = 10, p = p, max_round = 11)^cod_n10adj *
    prob_of_series_lasting_r_rounds(n = 11, p = p, max_round = 11)^cod_n11adj
}

@@

```{r}
#| label: cod-alternatives-to-p=0.5
cod_streaks_other_res <- tibble(
  p = c(0.5, cod_theoretical_p1, cod_theoretical_p2, cod_theoretical_p3),
  idx = 0L:3L,
  name = c('0. Naive', '1. Method of moments', '2. Maximum likelihood', '3. Minimum Chi-square')
) |> 
  mutate(
    series_streaks = map(
      p,
      ~summarize_cod_streaks(.x) |> 
        pluck('series') 
    ),
    chi_squ = map(
      series_streaks,
      ~.x |> 
        perform_series_chi_test() |> 
        select(statistic, p.value)
    )
  )

cod_streaks_other_res |>
  select(name, p, chi_squ) |>
  unnest(chi_squ) |>
  mutate(
    lab = generate_chi_label(statistic, p.value)
  ) |>
  select(
    `Method` = name,
    `$p$` = p,
    `$\\chi^2$ (p-value)` = lab
  ) |>
  knitr::kable()

cod_streaks_other_res |>
  filter(idx %in% c(0L, 2L)) |> 
  mutate(
    lab = sprintf('$p$ = %s', p)
  ) |>
  select(lab, series_streaks) |>
  unnest(series_streaks) |>
  select(lab, n_rounds, prop_expected) |>
  mutate(
    across(prop_expected, ~scales::percent(.x, accuracy = 0.1))
  ) |>
  pivot_wider(
    names_from = lab,
    values_from = prop_expected
  ) |>
  inner_join(
    cod_streaks_naive_res$series |>
      transmute(n_rounds, `$P_O(r)$` = scales::percent(prop_actual, accuracy = 0.1)),
    by = 'n_rounds'
  ) |>
  arrange(n_rounds) |>
  rename(`$r$` = n_rounds) |>
  knitr::kable()
```

Table \ref{tbl:mosteller-methods-results} shows the alternate values for the constant round win probability $p$ that I find when applying the three methods suggested by Mosteller [-@mosteller1952]. Each is approximately or equal to 0.575, and, when applying Equation \ref{eq:chi-squ}, each results in a $\chi^2$ value for which I cannot reject the constant probability null hypothesis.

```{=tex}
\begin{longtable}[]{@{}lrr@{}}
\caption{Alternate estimates of the constant probability \(p\) for winning a given round in a CoD SnD, applying the three methods suggested by Mosteller (1952), in addition to the naive \(p = 0.5\).}\label{tbl:mosteller-methods-results} \\
\toprule()
Method & \(p(r)\) & \(\chi^2\) (p-value) \\
\midrule()
\endfirsthead
\toprule()
Method & \(p(r)\) & \(\chi^2\) (p-value) \\
\midrule()
\endhead
0. Naive & 0.5000 & 16.0 (\textless=0.01) \\
1. Method of moments & 0.5725 & 3.6 (0.6) \\
2. Maximum likelihood & 0.5750 & 3.5 (0.62) \\
3. Minimum \(\chi^2\) & 0.5775 & 3.5 (0.62) \\
\bottomrule()
\end{longtable}
```
Table \ref{tbl:expected-series-lengths-alternative-ps} shows the new $P_E(r)$ when re-applying Equation \ref{eq:series-length} for the maximum likelihood estimate $p_2 = 0.575$.[^15] I observe that $P_E(r)$ is notably larger for $p_{2}$ when $r \in [6, 7]$, more closely matching $P_O(r)$. $p_{2}$ is also closer to $P_O(r)$ for $r \in [9, 10]$, although not for $r \in [8, 11]$.

[^15]: The method of moments and minimum $\chi^2$ estimates for $p$ are omitted simply because the results would be nearly identical to those for the maximum likelihood estimate of $p$ (since they are all $\approx 0.575$).

```{=tex}
\begin{longtable}[]{@{}rrrr@{}}
\caption{The probabilities that a best-of-11 series lasts $r$ rounds under the naive assumption $p_0 = 0.5$ and the maximum likelihood estimation $p_2 = 0.575$.}\label{tbl:expected-series-lengths-alternative-ps} \\
\toprule()
$r$ & $p_0(r)$ = 0.5 & $p_2(r)$ = 0.575 & $P_O(r)$ \\
\midrule()
\endhead
6 & 3.1\% & 4.2\% & 4.7\% \\
7 & 9.4\% & 11.2\% & 11.9\% \\
8 & 16.4\% & 17.8\% & 16.5\% \\
9 & 21.9\% & 21.8\% & 21.7\% \\
10 & 24.6\% & 23.0\% & 21.5\% \\
11 & 24.6\% & 22.0\% & 23.7\% \\
\bottomrule()
\end{longtable}
```
It is fair to conclude that the constant round win probability assumption can be valid in CoD SnD series with the appropriate choice of $p$ ($\approx 0.575$).

>>>>

There have been a handful of studies of the distribution of games played in a series of a professional sport. Most assume a constant probability $\phi$ of a given team winning a game in the series, regardless of the series state. Mosteller [-@mosteller1952] observed that the American League had dominated the National League in Major League Baseball's (MLB) World Series matchups, implying that matchups should not modeled with $\phi = 0.5$. Mosteller proposed three approaches for identifying the optimal constant probability value of the stronger team in the World Series, finding $\phi \approx 0.65$. in each case: (1) solving for $p$ from the empirical average number of games won by the loser of the series, which he called the "method of moments" approach; (2) maximizing the likelihood that the sample would have been drawn from a population in which the probability of a team winning a game is constant across the series (i.e. maximum likelihood), and (3) minimizing the chi-square goodness of fit statistic for $\phi$.

Chance [-@chance2020] re-examines the constant probability notion in Major League Baseball's (MLB) World Series (1923--2018), the National Basketball Association's (NBA) Finals (1951--2018), and the National Hockey League's (NHL) Stanley Cup (1939--2018). Chance finds strong evidence against the null hypothesis of $\phi = 0.5$ in the MLB and NHL championship series when applying Mosteller's first and second methods. Chance goes on to outline a conditional probability framework (likelihood of winning a game given the series state) which can exactly explain the distribution of the number of games played.

@@

Studying both player streaks and team streaks in basketball, in both observational and controlled settings. Gilovich et al. [-@gilovich1985], henceforth GVT, do not find evidence for the hot hand phenomenon. However, Miller and Sanjurjo [-@miller2018], henceforth MS, refuted the conclusions of GVT, proving a framework for quantifying streak selection bias, which effectively works in the manner posited by the gambler's fallacy[^9]. Specifically, MS say that a "bias exists in a common measure of the conditional dependence of present outcomes on streaks of past outcomes in sequential data" implying that, under i.i.d. conditions, "the proportion of successes among the outcomes that immediately follow a streak of consecutive successes is expected to be strictly less than the underlying (conditional) probability of success". When applying streak selection bias to GVT.'s data, MS come to the opposite conclusions as GVT.

[^9]: Streak selection bias, "in conjunction with a quasi-Bayesian model of decision making under sample size neglect... provides a novel structural candidate explanation for the persistence of gambler's fallacy beliefs", per MS.

Despite the plethora of existing research on games played in a series and momentum in sports, these topics have yet to be investigated heavily in esports. Work has been done to examine in-round win probability in other FPS titles such as Counter-Strike [@xenopoulos2022] and Valorant [@derover2021], both of which are round-based like CoD SnD. However, research on round-level trends is sparse, perhaps because games like Counter-Strike and Valorant both have economic aspects that can create clear advantages on side in a given round, given how prior rounds played out.

Additionally, both Counter-Strike and Valorant have overtime rules and blocked offensive/defensive roles (i.e. playing either offense or defense for many consecutive rounds). On the other hand, teams in CoD SnD rotate sides every round, analogous to a 1-1-1-1-1-1-1 format for home advantage in best-of-seven series for professional sports like the MLB, NBA, and NHL.[^10] While theoretically one might be able to account for any kind of balanced rotation, such as a "blocky" on like a 5-5-1, the rotation of team sides every round is convenient, particularly for convincing ourselves that rounds could reasonably be modeled as i.i.d. Bernoulli trials.

[^10]: 1-1-1-1-1-1-1 is not used today in these leagues, but it was at least one in each league.

@@

The formula for the expected proportion of seriesn a best-of-$s$ format ($s = 11$ rounds for CoD SnD), ending in $r$ rounds ($r \leq s$), $\hat{\Phi}(r)$, for a constant round win probability $\phi$ ($0 \leq \phi \leq 1$) for one team[^11] is

[^11]: If $\phi > 0.5$, then one might say that this team is the better team (known in hindsight).

@@

```{=tex}
\begin{equation}\protect\hypertarget{eq:m}{}{
m = \frac{s + 1}{2}
}\label{eq:m}\end{equation}
```
```{=tex}
\begin{equation}\protect\hypertarget{eq:series-length}{}{
\hat{\Phi}(r) = \frac{(r - 1)!}{(m - 1)!(r - s)!}(\phi^{m}(1 - \phi)^{r - m} + \phi^{r - m}(1 - \phi)^m).
}\label{eq:series-length}\end{equation}
```
To evaluate the constant round win probability hypothesis--that is, that the expected and observed round win rates, $\hat{\Phi}(r)$ and $\Phi(r)$ respectively, are equal to one another--we can compute the chi-square goodness of fit statistic, $\chi^2$.

```{=tex}
\begin{equation}\protect\hypertarget{eq:chi-squ}{}{
\chi^2 = \sum^R \frac{(\Phi(r) - \hat{\Phi}(r))^2}{\hat{\Phi}(r)}
}\label{eq:chi-squ}\end{equation}
```
For CoD SnD, $r \in R = [6, 7, 8, 9, 10, 11]$.

@@

The hindsight Bernoulli round loss probability after a streak of $k$ round wins, $p^{l|kr}_0$, can be formulated symmetrically. We might consider these as a reference when considering the

Removing our knowledge of a streak, we might model the Bernoulli round win probability as

```{=tex}
\begin{equation}\protect\hypertarget{eq:pwr}{}{
  p^{w|r} = p(\text{win | r})=\{
    \begin{array}{cl}
    \frac{m}{r}, & \text{team wins series} \\
    \frac{r - m}{r}, & \text{team loses series}
    \end{array}.
  }
\label{eq:pwr}
\end{equation}
```
As shown by MS, we should expect the proportion of rounds wins immediately following a streak of $k$ rounds wins for a series lasting $r$ rounds, $\hat{P}^{w|kr}_{MS}$, to be strictly less than $p^{w|r}$.[^12]

[^12]: See Appendix E.1 of MS for details on the proof that follows.

@@

MS note that there does not exists a closed form representation for $E[\hat{P}_k(\mathbf{X}) | I_k(\mathbf{X}) \neq \emptyset]$ for $k > 1$. ($\hat{P}_k(\mathbf{X})$ is the analogue for $\hat{P}^{w|kr}_{MS}$ in my notation for CoD SnD.) Figure 1 shows how the expected value varies as a function of the total number of trials $n$, given $p$ and $k > 1$.[^13]

[^13]: I've adapted the code from Vafa [-@vafa2017], which implements MS's framework [-@miller2018].

```{r}
#| label: npkd
## http://keyonvafa.com/hot-hand/
get_post_streak_prob <- function(n, k, p = 0.5) {
  tosses <- rbinom(n, 1, p)
  runs <- rle(tosses)
  n_neg_after <- length(which(runs$values == 1 & runs$lengths >= k))
  n_pos_after <- sum(runs$lengths[which(runs$values == 1 & runs$lengths >= k)] - k)

# library(extrafont)
# library(ggplot2)
# set.seed(42)
# runs <- rerun(
#   1000,
#   crossing(
#     n = 1:50,
#     k = 2:4,
#     p = 0.5
#   ) |>
#     mutate(
#       next_p = pmap_dbl(list(n, k, p), ~simulate_post_streak_prob(sims = 1000, n = ..1, k = ..2, p = ..3))
#     )
# ) |>
#   reduce(bind_rows)
# 
# agg_runs <- runs |>
#   filter(!is.nan(next_p)) |>
#   group_by(p, k, n) |>
#   summarize(
#     across(next_p, mean)
#   ) |>
#   ungroup() |>
#   arrange(p, k)
# 
# extrafont::loadfonts(quiet = TRUE)
# theme_set(theme_classic())
# theme_update(
#   text = element_text(family = 'CMU Typewriter Text')
# )
# p <- agg_runs |> 
#   arrange(p, k, n) |> 
#   mutate(
#     across(k, factor)
#   ) |> 
#   ggplot() +
#   aes(x = n, y = next_p, color = k, group = k) +
#   guides(
#     color = guide_legend(
#       title.hjust = 0.5,
#       nrow = 1, override.aes = list(size = 3)
#     )
#   ) +
#   geom_line() +
#   geom_hline(aes(yintercept = p)) +
#   labs(x = 'n', y = 'p') +
#   theme(
#     legend.position = c(0.5, 0.8)
#   )
# 
# ggsave(
#   p, 
#   filename = 'paper/images/npkd.png', 
#   width = 5.5, 
#   height = 3
# )
```

```{=tex}
\begin{figure}
\centering
\includegraphics{images/npkd.png}
\caption{The expected value of the proportion of successes on trials that immediately follow $k$ consecutive successes, $E[\hat{P}_k(\mathbf{X}) | I_k(\mathbf{X}) \neq \emptyset]$, for Bernoulli trial success $p = 0.5$ and streak lengths $k \in [2, 3, 4]$, as a function of the total number of trials $n.$}
\end{figure}
```
My context is fundamentally different from that of GVT and MS, both of whom focus on longitudinal data in controlled settings. GVT also perform statistical tests on shots from players in live games, i.e. "observational" data, but they note that their findings are likely affected player shot selection in the face of defensive strategy by the opposing team. The number of trials is fixed in their experimental designs, but in CoD SnD, the number of rounds played is determined as a function of the max number of possible rounds ($s$) and whether or not the team wins the series, as shown in Equation \ref{eq:pwr}. The Bernoulli trial success probability, i.e. a round win in CoD SnD, is not independent of the opponent.

Consequently, a statistical test of the difference in $\hat{P}^{w|kr}$ and $\hat{P}^{l|kr}$, as performed by GVT and MS to evaluate their respective hypothesis regarding $E[\hat{P}_k(\mathbf{X}) | I_k(\mathbf{X}) \neq \emptyset]$ (Theorem 1), is not appropriate.

Let us consider another form of the expected proportion of rounds won immediately after a streak of $k$ round wins in a best-of-$s$ series given the length of the series ($r$ rounds), the "hindsight" proportion $\hat{P}^{w|kr}_0$. The Bernoulli round win probability underlying $\hat{P}^{w|kr}_0$ is

~~When considering momentum in consecutive rounds wins, I use the descriptor "hindsight" because I account for the length of the series (~~$r$ rounds).

```{=tex}
\begin{equation}\protect\hypertarget{eq:pwkr}{}{
  p_0^{w|kr} = p_0(\text{win | k, r})=\{
    \begin{array}{cl}
    \frac{m - k}{i - k}, & \text{team wins series} \\
    \frac{s - m - k}{i - k}, & \text{team loses series}
    \end{array}.
  }
\label{eq:pwkr}
\end{equation}
```
A binomial test for the observed proportion of round wins following streaks of $k$ round wins in a series lasting $r$ rounds, $P^{w|kr}$, assuming $p_0^{w|kr}$ as the probability of success, can be used to verify that $p_0^{w|kr}$ is reasonable as a reference.

A proportion test of $\hat{P}^{w|kr}_0$ and $\hat{P}^{w|kr}_{MS}$ can tell us whether the momentum exists in CoD SnD. A null result means we cannot say that there is a difference between the hindsight expected proportion $\hat{P}^{w|kr}_0$ and the expected proportion adjusted for streak-selection bias, $\hat{P}^{w|kr}_0$.

@@

First, I investigate the constant probability assumption and the distribution of rounds played in a series. Chance's [-@chance2020] work is closely related to mine, and, in fact, provides a guide for this investigation. Afterwards, I investigate post-streak win rates and the context of momentum, leveraging MS's [-@miller2018] framework for streak selection bias.

@@

Using Equation \ref{eq:chi-squ}, I find that $\chi^2 = 16.0$ (p-value of 0.0068) for the naive constant round win probability $\phi_0 = 0.5$. Thus, I can comfortably reject the constant probability null hypothesis for $\phi_0 = 0.5$, even at a confidence level of $\alpha = 0.01$.

```{r}
#| label: cod_series_outcome_prop
.max_round <- 11
.cutoff <- 6
prob_of_series_lasting_r_rounds <- function(r, p = 0.5, s) {
  m <- as.integer((s + 1) / 2)
  (factorial(r - 1) / (factorial(m - 1) * factorial(r - m))) * (p^m * (1 - p)^(r - m) + p^(r - m) * (1 - p)^m)
}

@@

  expected_round_streaks <- tibble(
    n_rounds = .cutoff:.max_round
  ) |> 
    mutate(
      series_prop = map_dbl(n_rounds, ~prob_of_series_lasting_r_rounds(.x, s = .max_round, p = !!p)),
      n_expected_series_streaks = map_dbl(n_rounds, ~expected_series_streaks_of_outcomes(.cutoff, .x - .cutoff))
    ) |> 
    transmute(
      n_rounds,
      series_prop,
      prop = series_prop / n_expected_series_streaks
    )

@@

Table \ref{tbl:cod-prob-series-lasting-r-rounds} shows the expected series lasting $r$ rounds, $\hat{\Phi}_0(r)$ (under the assumption $\phi_0 = 0.5$), for $s = 11$, along with the observed proportions, $\Phi(r)$, in CoD SnD.

```{=tex}
\begin{longtable}{rrrr}
\caption{The expected proportion of series lasting $r$ rounds ($\hat{\Phi}_0(r)$) in a best-of-11 format, wehre $r \in R = [6, 7, 8, 9, 10, 11]$ under the assumption that each team has a constant round win probability $\phi_0 = 0.5$. The observed frequencies for CoD SnD are shown as a count $N(r)$ and as a proportion $\Phi(r)$ of all series ($\sum^R N(r)$).}\label{tbl:cod-prob-series-lasting-r-rounds} \\
\toprule
$r$ & $\hat{\Phi}_0(r)$ & $\Phi(r)$ & $N(r)$ \\ 
\midrule
6 & $3.1\%$ & $4.7\%$ & 40 \\ 
7 & $9.4\%$ & $11.9\%$ & 101 \\ 
8 & $16.4\%$ & $16.5\%$ & 141 \\ 
9 & $21.9\%$ & $21.7\%$ & 185 \\ 
10 & $24.6\%$ & $21.5\%$ & 183 \\ 
11 & $24.6\%$ & $23.7\%$ & 202 \\ 
\bottomrule
\end{longtable}
```
```{r}
#| label: cod_series_outcomes_naive_chi
generate_chi_label <- function(statistic, p.value) {
  sprintf('%.1f (%s)', statistic, ifelse(p.value <= 0.01, '<=0.01', as.character(round(p.value, 2))))
}

@@

```{r}
#| label: cod-alternative-to-p=0.5-method-1
## p. 365 on https://math.mit.edu/classes/18.095/2016IAP/lec9/Sports_Mosteller1952_WorldSeries.pdf
cod_prob_of_series_lasting_r_rounds <- function(r, p) {
  prob_of_series_lasting_r_rounds(r = r, p = p, s = 11)
}

theoretical_cod_series_length <- function(p) {
  6 * cod_prob_of_series_lasting_r_rounds(r = 6, p = p) +
    7 * cod_prob_of_series_lasting_r_rounds(r = 7, p = p) +
    8 * cod_prob_of_series_lasting_r_rounds(r = 8, p = p) +
    9 * cod_prob_of_series_lasting_r_rounds(r = 9, p = p) +
    10 * cod_prob_of_series_lasting_r_rounds(r = 10, p = p) +
    11 * cod_prob_of_series_lasting_r_rounds(r = 11, p = p)
}

cod_rounds_per_series <- cod_streaks_naive_res$series |>
  summarize(
    actual = sum(n_rounds * prop_actual),
    expected = sum(n_rounds * prop_expected)
  )

@@

maximize_cod_series_p <- function(p) {
  cod_prob_of_series_lasting_r_rounds(r = 6, p = p)^cod_n6adj *
    cod_prob_of_series_lasting_r_rounds(r = 7, p = p)^cod_n7adj *
    cod_prob_of_series_lasting_r_rounds(r = 8, p = p)^cod_n8adj *
    cod_prob_of_series_lasting_r_rounds(r = 9, p = p)^cod_n9adj *
    cod_prob_of_series_lasting_r_rounds(r = 10, p = p)^cod_n10adj *
    cod_prob_of_series_lasting_r_rounds(r = 11, p = p)^cod_n11adj
}

@@

```{r}
#| label: cod-alternatives-to-p=0.5
cod_streaks_other_res <- tibble(
  p = c(0.5, cod_theoretical_p1, cod_theoretical_p2, cod_theoretical_p3),
  idx = 0L:3L,
  name = c('0. Naive', '1. Method of moments', '2. Maximum likelihood', '3. Minimum Chi-square')
) |> 
  mutate(
    series_streaks = map(
      p,
      ~summarize_cod_streaks(.x) |> 
        pluck('series') 
    ),
    chi_squ = map(
      series_streaks,
      ~.x |> 
        perform_series_chi_test() |> 
        select(statistic, p.value)
    )
  )
```

Table \ref{tbl:mosteller-methods-results} shows the alternate values for the constant round win probability that I find when applying the three methods suggested by Mosteller [-@mosteller1952]. Each is approximately or equal to 0.575, and, when applying Equation \ref{eq:chi-squ}, each results in a $\chi^2$ value for which I cannot reject the constant probability null hypothesis.

```{=tex}
\begin{longtable}[]{@{}lrr@{}}
\caption{Alternate estimates of the constant probability ($\phi$) for winning a given round in a CoD SnD, applying the three methods suggested by Mosteller (1952), in addition to the naive ($\phi_0 = 0.5$).}\label{tbl:mosteller-methods-results} \\
\toprule()
Method & $\phi$ & $\chi^2$ (p-value) \\
\midrule()
\endfirsthead
\toprule()
Method & $\phi(r)$ & $\chi^2$ (p-value) \\
\midrule()
\endhead
0. Naive & 0.5000 & 16.0 (\textless=0.01) \\
1. Method of moments & 0.5725 & 3.6 (0.6) \\
2. Maximum likelihood & 0.5750 & 3.5 (0.62) \\
3. Minimum ($\chi^2$) & 0.5775 & 3.5 (0.62) \\
\bottomrule()
\end{longtable}
```
Table \ref{tbl:expected-series-lengths-alternative-ps} shows the new $\hat{\Phi}(r)$ when re-applying Equation \ref{eq:series-length} for the maximum likelihood estimate $\phi_2 = 0.575$, resulting in a new set of expected proportions of series lasting $r$ rounds $\hat{\Phi}_2(r)$.[^14] I observe that $\hat{\Phi}_2(r)$ is larger than $\hat{\Phi}_0(r)$ for $r \in [6, 7]$, more closely matching $\Phi(r)$. $\hat{\Phi}_2(r)$ is also closer to the observed $\Phi(r)$ for $r \in [9, 10]$, although not for $r \in [8, 11]$.

[^14]: The method of moments and minimum $\chi^2$ estimates for $\phi$ are omitted simply because the results would be nearly identical to those for the maximum likelihood estimate of $\phi$ (since they are all $\approx 0.575$).

```{=tex}
\begin{longtable}[]{@{}rrrr@{}}
\caption{The observed proportion of series $\Phi(r)$ ending in $r$ rounds in CoD SnD's best-of-11 format, compared to the expected proportion $\hat{\Phi}_0(r)$ the naive assumption $\phi_0 = 0.5$ and the expected proportion $\hat{\Phi}_2(r)$ under the maximum likelihood estimate $\phi_2 = 0.575$ for constant round win probability.}\label{tbl:expected-series-lengths-alternative-ps} \\
\toprule()
$r$ & $\hat{\Phi}_0(r)$ = 0.5 & $\hat{\Phi}_2(r)$ = 0.575 & $\Phi(r)$ \\
\midrule()
\endhead
6 & 3.1\% & 4.2\% & 4.7\% \\
7 & 9.4\% & 11.2\% & 11.9\% \\
8 & 16.4\% & 17.8\% & 16.5\% \\
9 & 21.9\% & 21.8\% & 21.7\% \\
10 & 24.6\% & 23.0\% & 21.5\% \\
11 & 24.6\% & 22.0\% & 23.7\% \\
\bottomrule()
\end{longtable}
```
Observing that $\hat{\Phi}_2(r)$ reasonably matches $\Phi(r)$ (especially in comparison to $\hat{\Phi}_0(r)$), along with the null hypothesis rejection shown in Table \ref{tbl:mosteller-methods-results}, it is fair to conclude that the constant round win probability assumption can be valid in CoD SnD series with the appropriate choice of $\phi$ ($\approx 0.575$).
